---
title: "Motor Trend Magazine"
author: "Rich Robinson"
date: "Saturday, June 13, 2015"
output:
  html_document:
    css: ~/R/win-library/3.1/rmarkdown/rmd/h/bootstrap-3.3.1/css/bootstrap.css
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    self_contained: no
    theme: null
---
```{r global_opts, include=FALSE}
knitr::opts_chunk$set(fig.align='center', fig.path='figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
## required libraries
library(datasets)
library(dplyr)
library(ggplot2)
library(GGally)

```

### Executive Summary

This analysis explores the relationship between the transmission type of cars and the affect this has on fuel efficiency, measured in Miles per gallon (MPG). In particular the analysis needs to address the following points:

* Is an automatic or manual transmission better for MPG?
* Quantify the difference in MPG between manual and automatic transmissions

To answer the above questions the analysis will first look into the basic statistical properties of the data using mean and standard deviation to gain a general idea of trends. Followering this a more detailed statitical analysis will include the fitting of linear models which can be used to predict future values within a level of confidence.

### Exploring the Data Set

We are looking at the *mtcars* dataset provided in `R`, so let's load it and have a look.

```{r load_dat}
data(mtcars)
head(mtcars)
```
We can see from the above table there is much information in each record, but for our analysis we are only interested in the `mpg` and `am` columns. The `am` data is a boolean value signifying if the transmission type is manual (1) or automatic (0). For ease of later analysis, we shall convert this in a factor variable.
```{r data_exp}
dat <- mtcars %>% select(mpg, am)

## adding a new column with am variable as a factor
dat <- dat %>% mutate(trans = factor(am, labels = c("automatic", "manual")))

g1 <- ggplot(dat, aes(x = trans, y = mpg)) + geom_point(color = dat$am+2)
g1 <- g1 + labs(title = "MPG against Transmission Type") + labs(x = "Transmission Type") + labs(y = "Miles per Gallon (MPG)")

## calculating some basic stats
stats <- dat %>% select(mpg, trans) %>% group_by(trans) %>% summarise_each(funs(mean, sd))
stats

## adding the stats as reference points on plot
g1 <- g1 + geom_point(data = stats, aes(x = trans, y = mean), color = "blue", size = 3)

```

The plot in appendix A shows the recorded MPG by transmission type. By observation it can be seen that for cars with automatic transmission the MPG is generally lower than that of manual cars. Also the range of MPG values for automatic cars is much smaller (more concentrated) than that of manual cars. The blue point for each transmission type represents the mean value of the data. The mean MPG for cars with automatic transmission is `r round(stats$mean[1], 3)` MPG and the mean for manual cars is `r round(stats$mean[2], 3)` MPG, which confirms the trend of higher MPG for manual cars. To qualtify the spread of the data we can look at the standard deviation from the mean. The standard deviation of automatic cars is `r round(stats$sd[1], 3)` MPG which is lower than that for manual cars of `r round(stats$sd[2], 3)` MPG, which indicates a range of values closer to the mean.

### Fitting a Linear Regression Model

It makes no sense to create a linear regression of MPG and transmission type as the predictor is a discrete quantity. A linear regression line would draw between the two discrete values, an area which is undefined. If we treat the transmission type as two separate groups, the best linear model will be a constant, namely the mean of the group, $\beta_0 = \mu$, and $\beta_1$ (the slope) will be undefined. Thus we must look for more significant predictors to MPG.

### Attribute Correlations

It is logical that more attributes will affect MPG than just transmission type. Now lets look for strong correlations between other attributes and MPG using the `corr()` function.

```{r corrs}
## calculating correlations of the data
foo <- cor(mtcars, mtcars$mpg)
corrs <- data.frame(attr = rownames(foo), corr = foo[,1])
## arranging into order of significance by ABS value to preserve relationship (positive or negative)
corrs <- arrange(corrs, desc(abs(corr)))
corrs
```

The output above shows that many attributes have a stronger affect on MPG than transmission type (am), namely weight, cylinders, displacement, horse power. This would suggest that a multivariate regression model would better predict the possible MPG of a vehicle, with predictors of weight, cylinders, displacement and horse power. Lets look into this further. (See appendix B for the code for the above calculation and a pairs plot for illustration)

### Appendix A - Code & Plot of MPG vs Transmission
```{r data_exp, echo=TRUE, eval=FALSE}
```
```{r app_a, fig.cap="Figure 1"}
g1
```

### Appendix B - Code Calculating Correlations between Attributes
```{r corrs, echo=TRUE, eval=FALSE}
```
```{r pairs_corr, echo=TRUE, fig.cap="Figure 2"}
## selecting column numbers
col_nums <- as.numeric(order(desc(abs(foo))))
## creating pairs plot
ggpairs(select(mtcars, col_nums[1:5]), title = "Pairs plot of most Significant Attributes towards MPG")
```